---
title: "Introduction to R Programming"
author: "Yulia Iakovleva"
format: html
editor: visual
---

# Data Structures in R

## Vectors

Vectors come in two flavours: **atomic vectors** and **lists**. They differ in terms of their elements' types: for atomic vectors, all elements must have the same type; for lists, elements can have different types.

### Atomic Vectors

-   A vector is the most basic data structure in R.

-   More complex structures (matrix, list, factor, data frame) are based on vectors.

-   Vectorization is a key concept of the language.

### Definition of Atomic Vector

-   A vector is an indexed set of data of the same type.

-   R does not distinguish between scalar and vector values; that is, a scalar is a vector of length 1.

-   Indexing of vectors starts from one (not zero!).

-   To create a vector, you can use the `vector` function.

```{r}
x <- vector(length = 2)
x[1] <- 5
x[2] <- 8
x
```

### How to Create Atomic Vector

#### Combine Function

Instead of vector you can use combine `c` function:

```{r}
x <- c(5, 8)
x
```

The function `c` can be nested and accept variables:

```{r}
y <- c(x, 1, c(3, 4), x, NA)
y
```

#### Colon Operator

Colon operator creates the series of numbers in sequence for a vector (numerical sequences with a step of 1 only).

```{r}
5:9
3:-1
```

#### Function `seq`

For more complicated sequences use `seq` function:

```{r}
seq(1, 2, by = 0.25)
seq(3, 4, length = 6)
```

#### Function `rep`

To replicate or repeat elements in a vector or a sequence.

```{r}
rep(1:3, times = 3)
rep(1:3, each = 2)
rep(1:3, length = 4)
```

### Types of Atomic Vectors in R

All elements of atomic vectors must have the same type:

-   Logical

-   Integer

-   Numeric (integer, double)

-   Character

-   ...

#### Get Vector Type

To determine the type of a vector, there are functions `typeof` and `is.*` (where `*` represents any of the types).

```{r}
text <- c("Lorem", "ipsum", "dolor", "sit", "amet")
typeof(text)
is.numeric(text)
```

#### Factors

Factors are data structures that are implemented to categorize the data or represent categorical data and store it on multiple levels.

```{r}
x <- c("Man", "Male", "Man", "Lady", "Female")
## Map from 4 different values to only two levels:
(xf <- factor(x, levels = c("Male", "Man" , "Lady",   "Female"),
                 labels = c("Male", "Male", "Female", "Female")))
xf
```

### Length of Vector

```{r}
x <- 1:100
length(x)
```

### Named Vectors

Elements of vectors could be named:

```{r}
a <- c(uno = 1, dos = 2, "universal answer" = 42, 99)
names(a)
```

## Operations on Vectors

You can do both arithmetical and logical. Operations of vectors are performed member-by-member, i.e., *memberwise*.

```{r}
1:3 + c(-1, 2, 0)
```

```{r}
1:3 * c(-1, 2, 0)
```

```{r}
c(TRUE, TRUE, TRUE) & c(0, 1, 999)
```

### Exercise

The results are different. Why?

```{r}
c(T, F) + c(0, 1)
c(T, F) & c(0, 1)
```

## Recycling

When applying an operation to two vectors that requires them to be the same length, R automatically *recycles*, or repeats, the shorter one, until it is long enough to match the longer one.

```{r}
vec1 <- 1
vec2 <- c(1, 2, 3)

vec1 + vec2
```

## Accessing Vectors

Get elements of vector from 2d to 5th:

```{r}
x <- 1:10
x[2:5]
```

Filtering:

```{r}
x[x > 4]
```

## Lists

-   A list is an indexed structure.

-   They are typically used to store data of **varying lengths** or **different types**.

-   Lists are created using the **`list`** function.

How to create list:

```{r}
list(1:10)
```

It might be nested or recursive:

```{r}
list(list(1, 2, 3), c("a", "b"), "ATGCACA")
```

As vectors list could be named:

```{r}
l1 <- list(name = "john", salary = 1000)
l2 <- list(has_car = TRUE, car = "lamborghini")
```

List could be also concatenated:

```{r}
l3 <- c(l1, l2)
```

### Accessing List

#### Subset List

```{r}
l3[1]
```

#### Get the Single Element

```{r}
l3[[1]]
l3[["name"]]
```

#### Dollar Sign Operator

The Dollar `$` sign in R is a special operator that is used to access the List or DataFrame. Similar to bracket `[]` notation, you can use `$` sign to access, add, update and delete variables from list and columns from DataFrame.

```{r}
l3$name
```

#### Summary

Single Brackets

-   Follow vector indexing rules

-   Returned value is sublist

Double Brackets

-   Index number or its full name

-   Returned value is list element (atomic vector)

Dollar Sign

-   We can acess by partial name of the element

-   Returned value is list element (atomic vector)

### Replace or Add Element to List

To add new element with "company" name and "Big Industry" value:

```{r}
l3$company <- "Big Undustry"
l3
```

The same you can do to replace existing value in list.

## Apply Family Functions

The `apply()` family pertains to the R base package and is populated with functions to manipulate slices of data from matrices, arrays, lists and dataframes in a repetitive way. These functions allow crossing the data in a number of ways and avoid explicit use of loop constructs. They act on an input list, matrix or array and apply a named function with one or several optional arguments.

Example:

```{r}
l <- list(A = c(56, 12, 57, 24), B = c(89, 12, 64, 18, 65, 76))

lapply(l,                           # List
       quantile,                    # Applied function
       probs = c(0.25, 0.5, 0.75))  # Additional argument of the quantile function
```

However, it is considered legacy functionality and should not be used for new code. Instead, it is recommended to use the `purrr` package for all looping in R. The `purrr` package provides a consistent syntax for working with functions that take multiple inputs and outputs, making it easier to write and read code. Additionally, purrr offers a suite of functions that are optimized for modern computing environments, making it faster and more efficient than the Apply Family. Overall, while the R Apply Family may still work for some applications, it is advisable to use the purrr package for all new R projects.

## For Loop vs Vectorization

In R, for loops are generally slower than vectorized operations because of the way R is designed and optimized for vectorized operations. Here are some key reasons why for loops can be slower compared to vectorization in R:

-   **Internal Optimization**: R is an interpreted language, and it has been optimized for vectorized operations.

-   **Function Call Overhead**: In R, function calls can be relatively expensive in terms of performance. When you use a for loop to iterate over elements and perform a simple operation on each element, you are making multiple function calls within the loop (e.g., accessing and modifying elements). This overhead can slow down the execution of the loop.

-   **Memory Allocation**: Vectorized operations in R often require less memory allocation compared to for loops. When you use a for loop to build a result iteratively, R may need to allocate memory for each iteration, which can be less efficient than allocating memory for an entire vector in one go.

-   **Ease of Parallelism**: Vectorized operations in R are more conducive to parallelism. Many modern CPUs have multiple cores, and R can take advantage of this by parallelizing vectorized operations. In contrast, for loops are typically executed sequentially, making it harder to utilize multiple CPU cores effectively.

-   **Optimized Libraries**: R has many optimized libraries and packages that are designed to work with vectors and matrices efficiently. These libraries often provide vectorized functions that can perform complex operations on large datasets much faster than equivalent for loops.

```{r}
v <- 1:1e5
system.time({
  x <- 0
  for (i in v) x[i] <- sqrt(v[i])
})
```

```{r}
system.time({
  y <- sqrt(v)
})
identical(x, y)
```

⚠️ Don't use for loop on large data. Use vectorization approach instead.
