---
title: "Introduction to R Programming"
author: "Yulia Iakovleva"
format: html
editor: visual
---

# Data Frames in R

## What is Data Frame?

-   Two-dimensional table with data (like Excel Spreadsheet or SQL table).

-   The standard way of storing data in the "observations/variables" format: rows correspond to observations, columns correspond to variables.

-   Data frame inherits properties from both a matrix (rectangular shape) and a list (variables can be of different types).

More about R Data Frames see [here](https://www.geeksforgeeks.org/r-data-frames/).

## How to Create

```{r}
df <- data.frame(
  x = 1:4,
  y = LETTERS[1:4],
  z = c(T, F))

df
```

## How to Explore

In R there are many built-in data sets that we can use for our exercises.

To explore the variety of data set, you can call the `data()` function.

```{r}
data()
```

Today we will work with `mtcars` data set. To get more about it, get help:

```{r}
?mtcars
```

Get the first part of data (by default the first 6 rows):

```{r}
head(mtcars)
```

Get data set structure

```{r}
str(mtcars)
```

Get row names and column names:

```{r}
rownames(mtcars)
colnames(mtcars)
```

Dimensions

```{r}
# Get the number of columns
rownames(mtcars)
# Get the numbe of raws
colnames(mtcars)
# Get dimentions
dim(mtcars)
```

## Data Frame Manipulations

### Acceding Data

Accessing data frames is pretty similar to accessing atomic vectors or list.

```{r}
mtcars[1:10, 1:5]
```

### Filtering

By conditions

```{r}
mtcars[mtcars$mpg > 20, ]
```

It's possible to include multiple conditions:

```{r}
mtcars[(mtcars$cyl > 4) & (mtcars$hp > 200), ]
```

You can also use `subset` function to filter your data frame or select specific columns:

```{r}
subset(mtcars, gear == 4, select = c(mpg, cyl, gear))
```

### Combining Data Frames

Functions `rbind` and `cbind` built-in functions.

#### Combine Rows

```{r}
rbind(
  df,
  data.frame(
    x = 5:6,
    y = c("K", "Z"),
    z = TRUE,
    row.names = c("Kappa", "Zulu")
  )
)
```

#### Combine Columns

```{r}
cbind(
  df,
  data.frame(
    season = c("Summer", "Autumn", "Winter", "Spring"),
    temp = c(20, 5, -10, 5)
  )
)
```

### Merge by Column

```{r}
merge(
  df,
  data.frame(
    x = c(3, 2, 6, 1),
    salary = c(100, 1000, 300, 500)
  ),
  by = "x" # specify by which column do you want to merge
)
```

## Data Pre-processing Workflow

### Workflow Stages

-   Import into a data frame

-   Cleaning values, checking types

-   Working with strings: names, character variables, factors

-   Missing values: identification, handling method

-   Manipulating variables: transformation, creation, deletion

-   Calculating descriptive statistics

-   Data visualization

-   Export

### Reading Data

Main function is [read.table](https://www.geeksforgeeks.org/reading-contents-of-a-text-file-in-r-programming-read-table-function/).

-   **file** file name

-   **header** presence or absence of a header in the first row

-   **sep** value separator (usually we have CSV or TSV)

-   **dec** decimal point

-   **quote** characters denoting quotes (for string type)

-   **na.strings** strings encoding missing values

-   **colClasses** column types (for performance and specifying type: character, factor, date/time, etc.)

-   **comment.char** character denoting a comment

-   **skip** number of lines to skip from the beginning of the file

See more for details:

```{r}
?read.table
```

Functions `read.csv`, `read.csv2`, `read.delim`, and `read.delim2` are wrappers around `read.table` with default settings.

Summary on how to read different types of data:

-   For Excel Spreadsheets with [readxl](https://readxl.tidyverse.org/).

-   Read CSV / TSV Files with [read.csv](https://www.geeksforgeeks.org/read-contents-of-a-csv-file-in-r-programming-read-csv-function/).

-   For more complex text data [read.table](https://www.geeksforgeeks.org/reading-contents-of-a-text-file-in-r-programming-read-table-function/).

### **Cleaning Values, Type Checking**

Variable types that can easily lead to errors during import:

-   Numeric types become character due to missing values, marked not as `NA`: `na.strings = c("NA", "Not Available", "Missing")`

-   Numeric types become character due to incorrectly specified separators or decimal points: `sep = ","`, `dec = "."`

-   Numeric types become character due to quotes, accompanying text, or comments: `quote`, `comment.char`, `skip`

-   String types become factors or character instead of necessary type of vector: `as.character`, `as.factor`

-   Date/time types remain character: `as.POSIXct`, `as.POSIXlt`, `as.Date`

Functions `str`, `summary`, `head`, and `tail` can help determine if everything is in order.

### **Working with Variables**

-   Functions `complete.cases` and `na.omit` for removing observations with missing values: `df[complete.cases(df), ]` or `na.omit(df)`

    For example:

```{r}
df <- data.frame(
  x = c(1,2,3),
  y = LETTERS[1:3],
  z = c('yes', 'no', NA)
)

# Get rows that contains NAs
df[!complete.cases(df), ]
# Remove NAs
na.omit(df)
```

-   Replacing NA with certain values can be potentially risky:

    -   Filling with the mean can introduce bias into the data.

    -   Filling with zeros is generally incorrect in most cases.

### **Export**

-   `write.table`, `write.csv`, and `write.csv2` are almost identical to the import functions.

-   If the data array is large, it's best to separate the data pre-processing stage into:

    -   a separate .R file

    -   a separate text file with pre-processed ("clean") data.

## Write Readable Code: dplyr

To write a readable code it is highly recommended to have a look at dplyr R package. It is a grammar of data manipulation, providing a consistent set of verbs that help you solve the most common data manipulation challenges.

For example:

```{r}
library(dplyr)

mtcars %>%
  # Select columns by names
  select(mpg, cyl, gear) %>%
  # Filter rows by condition
  filter(mpg > 20) %>%
  # Create new column with 'new_gear' name
  mutate(new_column = c(T)) %>%
  # Rename new_gear 
  rename(is_filtered = new_column) %>%
  # For each group of gear get the mean mpg value
  group_by(gear) %>%
  summarise(mean(mpg))
```

## References

-   [Advanced R](https://adv-r.hadley.nz/) book by Hadley Wickham.
